Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-27 analysis/paper/paper.Rmd
updated the isotope plot

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994].

@@

# plot the carbon isotopes 
isotope_C16_C18 <-
  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
       aes(`13C C16:0`,`13C C18:0`)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  #ggrepel::geom_text_repel(aes(label = C13)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
  xlim(-33,-20) +
  ylim(-33,-20) +
  #coord_fixed(ratio = 1) +
  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
  geom_abline(intercept = -3.1, linetype = "dashed") +
  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))

big_delta_C16 <-
  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
       aes(`13C C16:0`, delta)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  ylim(-2.5, 2) +
  geom_hline(yintercept = -0.7, linetype = "dashed") +
  annotate("text", x = -22, y = -0.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -31, y = 1.8,
           label = "C3-diet") +
  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x = -22, y = 1.8,
           label = "C4-diet")

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in an earlier period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan. Therefore, it is more likely that deer could be the main resource of ruminant fats. The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings in the future.

check isotopic results for horse, deer and cattle

>>>>

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Evershed1994].

@@

# plot the carbon isotopes 
isotope_C16_C18 <-
  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
       aes(`13C C16:0`,`13C C18:0`)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  #ggrepel::geom_text_repel(aes(label = C13)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  scale_y_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  #coord_fixed(ratio = 1) +
  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
  geom_abline(intercept = -3.1, linetype = "dashed") +
  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))

big_delta_C16 <-
  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
       aes(`13C C16:0`, delta)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  ylim(-2.5, 1.5) +
  geom_hline(yintercept = -0.7, linetype = "dashed") +
  annotate("text", x = -31.5, y = -0.9,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -32, y = 1.3,
           label = "C3-diet") +
  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x = -21, y = 1.3,
           label = "C4-diet") +
  annotate("text", x = -22, y = -1.1,
           label = "Ruminant") +
  annotate("text", x = -21.5, y = -0.3,
           label = "Non-ruminant")

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to a later period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains from the sites, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in an earlier period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan in an early period. Therefore, it is more likely that deer could be the resource of ruminant fats. The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings.

check isotopic results for horse, deer
